---
title: "Transformation of quantitativ traits"
author: "AAZaidi"
date: "2022-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(data.table)
library(ggplot2)
library(MASS)
F = rprojroot::is_rstudio_project$make_fix_file()
```


Determining the right transformation to use for each lab. First, load in the lab data and look at their distributions.


```{r}

quants = fread(F("data/phenotype/PMBB_2.2_quant_08112022.pheno"))

#remove rlrmtCN (we know it's normal)
quants = quants[variable !="rlrmtCN"]

quants[, scaled_value := scale(value), by = "variable"]

ggplot(quants, aes(scaled_value))+
  geom_density()+
  facet_wrap(~variable, scales = "free")+
  theme_classic()+
  labs(x = "Phenotype (SD units)")

```

Some of the distributions (e.g. serum creatinine and CRP) are extremely skewed while others are not. There are also some really short people here.

```{r}

quants[variable == "Height" & scaled_value < -5]


```

These are clearly wrong. Let's remove these.

```{r}

quants[variable == "Height" & scaled_value < -5, value := NA]
quants[variable == "Height" & scaled_value < -5, scaled_value := NA]

#also remove negative values
quants[value < 0, value := NA]
quants[value < 0, scaled_value := NA]

#also remove values that are crazy high (e.g. > 7 SD)
quants = quants[scaled_value < 7 & scaled_value > -7]

quants = na.omit(quants)

ggplot(quants, aes(scaled_value))+
  geom_density()+
  facet_wrap(~variable, scales = "free")+
  theme_classic()+
  labs(x = "Phenotype (SD units)")

```

Let's try a log transformation to all traits.

```{r}

ggplot(quants[variable != "rlrmtCN"], aes(log(value+1)))+
  geom_density()+
  facet_wrap(~variable, scales = "free")+
  theme_classic()+
  labs(x = "Phenotype (Log transformed)")

```

Most of these look a lot better. Let's try the Box-Cox transformation. 

```{r}

mtcn = fread(F("data/phenotype/PMBB_plink4mat_combined.blood_residuals_07272022.pheno"))
pcs.eur = fread(F("data/gcta/EUR/PMBB-Release-2020-2.0_genetic_genotype.EUR.maf0.01.pca.header.eigenvec"))
demog = fread(F("data/phenotype/PMBB-Release-2020-2.2_phenotype_covariates.txt"), header = TRUE)

demog[is.na(Age_at_Enrollment) == FALSE, `:=`(Age, Age_at_Enrollment)]
demog[is.na(Age_at_Enrollment) == TRUE, `:=`(Age, Age_first_sample)]
demog = demog[Age > 0]

covar = demog[, .(PMBB_ID, Sex, Age)]
mtcn.eur = merge(mtcn, pcs.eur, by = c("FID", "IID"))
mtcn.eur = merge(mtcn.eur, covar, by.x = "IID", by.y = "PMBB_ID")

phenotypes = unique(quants$variable)
phenotypes = setdiff(phenotypes, "rlrmtCN")

quants.eur = merge(quants, mtcn.eur, by.x = "PMBB_ID", by.y = "IID")
pc.form = paste(paste("PC", seq(1,20,1), sep = ""), collapse = "+")

trans.labs = list()
lambdas = list()
for(i in 1:length(phenotypes)){
  x = phenotypes[i]
  dat = quants.eur[variable == x]
  dat[, y := value]
  if(min(dat$value) == 0){
    dat[, y := value + 1 ]
  }
  form = paste("y ~ Sex + poly(Age, 2, raw = TRUE)+", pc.form, sep = "")
  bt = boxcox(lm(data = dat, form), plotit = FALSE)
  lambda = bt$x[which.max(bt$y)]
  lambdas[[i]] = data.table(lambda = lambda)
  if(lambda == 0){
    dat[, tvalue := log(y)]
  }else{
    dat[, tvalue := (y^lambda - 1)/lambda]
  }
  trans.labs[[i]] = dat
}

names(trans.labs) = names(lambdas) = phenotypes
trans.labs = dplyr::bind_rows(trans.labs, .id = "Phenotype")

```

Plot the distributions

```{r}

ggplot(trans.labs, aes(tvalue))+
  geom_density()+
  theme_classic()+
  labs(x = "Phenotype (transformed)")+
  facet_wrap(~Phenotype, scales = "free")

```

these look better though some don't look like they will improve with a transformation. Output these to file. Print the lambdas so we can back-transform later. 


```{r}

#transform all traits (in both EUR and AFR) and save

trans.labs = list()
for(i in 1:length(phenotypes)){
  x = phenotypes[i]
  dat = quants[variable == x]
  dat[, y := value]
  if(min(dat$value) == 0){
    dat[, y := value + 1 ]
  }
  lambda = lambdas[[i]]
  if(lambda == 0){
    dat[, tvalue := log(y)]
  }else{
    dat[, tvalue := (y^lambda - 1)/lambda]
  }
  trans.labs[[i]] = dat
}

names(trans.labs) = phenotypes
trans.labs = dplyr::bind_rows(trans.labs, .id = "Phenotype")

ggplot(trans.labs, aes(tvalue))+
  geom_density()+
  theme_classic()+
  labs(x = "Phenotype (transformed)")+
  facet_wrap(~Phenotype, scales = "free")


fwrite(trans.labs[, .(PMBB_ID, Phenotype, tvalue)],
       F("data/phenotype/PMBB_2.2_quant_08122022.boxcox.pheno"), sep = "\t")

```


```{r}

lambdas = dplyr::bind_rows(lambdas, .id = "Lab")

fwrite(lambdas, F("output/tables/boxcox_lambdas_08302022.txt"), sep = "\t")

```





